Galaxy image statistical generator: how to be confident?

### 2024-09-23 paper/main.tex
GAN lightweight

It is usual to call $p_{\bm{\theta}_2}(\bm{x}|\bm{z})$ the \textit{decoder} part of the whole model while $p_{\bm{\phi}}(\bm{z}|\bm{x})$ is the \textit{encoder} part. The different VAE models differ by the distribution families used. The simplest choice may be commonly-used Gaussian multivariate distributions, and the centred isotropic with unit variance version ($\mathcal{N}(\bm{z},\bm{0},\bm{1}$) for the $\bm{z}$ prior (nb. $\bm{\theta}_1$ are missing is this scenario). Once the optimization is done, the \textit{decoder} is used as a $\bm{x}$ generator by feeding with $\bm{z}$ samples. It turns out that simple VAE tends to generate blurry images and may  suffer from the problem of \textit{posterior collapse} which motivate active ML researches \citep[e.g.][]{engel2018latent,Takida2022} and modified VAE versions to overcome this issue as for instance in \cite{Lanusse2021}. 
%
\subsection{Generative Adversarial Networks}
%
The vanilla GAN \citep{goodfellow2014generative} shares with VAE the target to optimize a \textit{generator} network ($\bm{G}$) to get $\bm{x}$ samples from a latent variable $\bm{z}$ such that  $\bm{x} = \bm{G}(\bm{z})$ with $\bm{z}$ \textit{prior} $\pi(\bm{z})$ is for instance a Gaussian distribution $\mathcal{N}(\bm{z},\bm{0},\bm{1})$. To do so, a second network ($\bm{D}$) acts as a \textit{discriminator} which aims to state if a sample $\bm{x}$ is from the model distribution or the data distribution. To reach that goal one solves the \texttt{min-max} problem
\begin{equation}
\min_{\bm{G}} \max_{\bm{D}}\left\{ \Esp_{\bm{x}\sim p_{data}(\bm{x})}[\log \bm{D}(\bm{x})] + \Esp_{\bm{z}\sim \pi(\bm{z})}[\log(1-\bm{D}(\bm{G}(\bm{z})))] \right\}
\end{equation} 
where $p_{data}$ is the data generating distribution. Both the  $\bm{G}$ and $\bm{D}$ are parametrized networks. Notice that at parameter of $\bm{G}$ fixed, the optimal \textit{discriminator} $\bm{D}^\ast$ has the following expression
\begin{equation}
\bm{D}^\ast(\bm{x}) = \frac{p_{data}(\bm{x})}{p_{data}(\bm{x}) + p_{\bm{G}}(\bm{x})}
\end{equation}
with $p_{\bm{G}}(\bm{x}) = \pi(\bm{z})|\det J_{\bm{G}} |^{-1}$ using the Jacobian of the \textit{generator}, leading to a reformulation of the \texttt{min-max} problem as followed
\begin{equation}
\min_G \left\{ \DKL{\left( p_{data} \left\| \frac{p_{data} + p_{\bm{G}}}{2}\right. \right) }  + \DKL{\left(p_{\bm{G}} \left\| \frac{p_{data} + p_{\bm{G}}}{2} \right. \right)} \right\}
\end{equation}
leading to the optimal solution $p_{\bm{G}} = p_{data}$ yielding a perfect matching of the data distribution.

GAN architectures have been used for galaxy generation as \citep[e.g.][]{Schawinski2017,Coccomini2021,Hemmati_2022} in conjunction of deblending and deconvolving tasks. 
%
\subsection{Normalizing Flows}
%
Looking at Equation \ref{eq-VAE-ELBO}, one notices that the optimal solution is reached in case of $\DKL(p_{\bm{\phi}}(\bm{z}|\bm{x})\| p(\bm{z}|\bm{x}))=0$ which is true iif $p_{\bm{\phi}}(\bm{z}|\bm{x}) = p(\bm{z}|\bm{x})$. But this situation cannot be true in general as typical $p_{\bm{\phi}}$ parametrizations involve for instance independent Gaussian distributions. The limited choices of the parametrization is one of the limitation of VAE. So, one motivates the use of normalizing flows (NF) to allow for more flexibility in the distribution family design to hope that the true posterior is member of such family \citep{Tabak2010, Tabak2013a, Rezende2015}.

@@

%
\subsection{Datasets and Models}
%
An answer to the previous question has been given in \cite{kadkhodaie2024generalization} and point out the size of the dataset in the passage from \textit{memorization} to \textit{generation} regimes. We propose to replay their experiment using first the same \texttt{U-Net} architecture \citep{2015arXiv150504597R}  as \textit{denoiser} consisting of a contracting part and an expending part that allows for multi-scale architecture with scale correlations. Details are found in \cite{kadkhodaie2024generalization}\footnote{Source code:  \url{https://github.com/LabForComputationalVision/memorization_generalization_in_diffusion_models}}. Second, for comparison purposes, we use the \texttt{Glow} flow-based model implemented\footnote{Source code: \url{https://github.com/rosinality/glow-pytorch}} in \texttt{PyTorch}  \citep{PyTorch2019}. The architecture has been presented on Figure \ref{fig-Glow-archi} and we use $K=32$ and $L=4$.

Concerning the training dataset, \cite{kadkhodaie2024generalization}  have used different datasets commonly-used in ML as \texttt{CelebA} \citep{Liu2015}, \texttt{LSUN} bedroom \citep{Yu2015} and  \texttt{CelebA HQ} \citep{KarrasALL18} datasets. All dataset images use  gray level single channel ranging in $[0,1]$.  The \texttt{LSUN}  images have been down sampled to $80\times 80$ as well as the \texttt{CelebA} images which also have been further down sampled to $32\times 32$ resolution, while \texttt{CelebA HQ}  have been down sampled to $40\times 40$ resolution.

We are more concerned with galaxy image generation, but as we need to train two models with up to $100,000$ images each, we have chosen to use the SDSS (Sloan Digital Sky Survey) dataset described in \citep{smith2021}\footnote{See \url{https://github.com/Smith42/astroddpm} to obtain the raw data.} with some modification on the preprocessing. First, all images are cropped about the target coordinate galaxy to a shape of $64\times 64$ pixels, and secondly from the $(z,g,r)$ channels we produce a single  channel in the range $[0,1]$ using \texttt{scipy} \citep{2020SciPy-NMeth} library\footnote{In particular the \texttt{make\_lupton\_rgb} function has been applied with $Q=8$ and $stretch=0.2$ setting}. Moreover, the gray levels are coded using 5-bit precision in the line of experiment described in \cite{Kingma2018}. With such dataset according to \cite{kadkhodaie2024generalization} results, we expect the transition between memorization to generalisation regimes to occur with a number of training images in the range $10^4-10^5$.

@@

Figure \ref{fig-Original-Glow-UNet-samples} presents some galaxy images extracted from the original dataset and generated by  diffusion-based and flow-based models trained with $N=10^5$ original images. 

@@

    \vspace{-0cm}
    \begin{subfigure}[t]{0.30\columnwidth}
    \begin{center}
        \includegraphics[width=0.8\textwidth]{fig-original_samples_gray.pdf}
        %\caption{xxx}
    \end{center}
    \end{subfigure}%
    \begin{subfigure}[t]{0.30\columnwidth}
    \begin{center}
        \includegraphics[width=0.8\textwidth]{fig-diffusion_samples_gray.pdf}
        %\caption{xxx}
    \end{center}
    \end{subfigure}%
    \begin{subfigure}[t]{0.30\columnwidth}
        \raisebox{0mm}{\includegraphics[width=0.8\textwidth]{fig-glow_samples_gray.pdf}}
        %\caption{xxx}
    \end{subfigure}
    \caption{Illustration of the generation of some samples from the diffusion-based (\texttt{U-Net} denoiser) and the \texttt{Glow} flow-based models trained with $N=10^5$ images. On the left some original images are also displayed for visual comparison.}
    \label{fig-Original-Glow-UNet-samples}

@@

	\caption{Corner plots \citep{corner2016} of morphological coefficients computed with $10^4$ images, first from the original dataset but \textit{never} used by any models (red, "test"), second from a diffusion-based model generation (blue, "diffusion") and from a flow-based model generation (green, "Glow") (both models used for Figure \ref{fig-Original-Glow-UNet-samples}). The figure has been inspired by \cite{HACKSTEIN2023100685}.}

@@

>>>>

It is usual to call $p_{\bm{\theta}_2}(\bm{x}|\bm{z})$ the \textit{decoder} part of the whole model while $p_{\bm{\phi}}(\bm{z}|\bm{x})$ is the \textit{encoder} part. The different VAE models differ by the distribution families used. The simplest choice may be commonly-used Gaussian multivariate distributions, and the centred isotropic with unit variance version ($\mathcal{N}(\bm{z},\bm{0},\bm{1}$) for the $\bm{z}$ prior (nb. $\bm{\theta}_1$ are missing is this scenario). Once the optimization is done, the \textit{decoder} is used as a $\bm{x}$ generator by feeding with $\bm{z}$ samples. It turns out that simple VAE tends to generate blurry images and may  suffer from the problem of \textit{posterior collapse} which motivate active ML researches \citep[e.g.][]{engel2018latent,Takida2022} and modified VAE versions to overcome this issue as for instance in \cite{Lanusse2021}. 
%
\subsection{Generative Adversarial Networks}
%
The vanilla GAN \citep{goodfellow2014generative} shares with VAE the target to optimize a \textit{generator} network ($\bm{G}$) to get $\bm{x}$ samples from a latent variable $\bm{z}$ such that  $\bm{x} = \bm{G}(\bm{z})$ with $\bm{z}$ \textit{prior} $\pi(\bm{z})$ is for instance a Gaussian distribution $\mathcal{N}(\bm{z},\bm{0},\bm{1})$. To do so, a second network ($\bm{D}$) acts as a \textit{discriminator} which aims to state if a sample $\bm{x}$ is from the model distribution or the data distribution. To reach that goal one solves the \texttt{min-max} problem
\begin{equation}
\min_{\bm{G}} \max_{\bm{D}}\left\{ \Esp_{\bm{x}\sim p_{data}(\bm{x})}[\log \bm{D}(\bm{x})] + \Esp_{\bm{z}\sim \pi(\bm{z})}[\log(1-\bm{D}(\bm{G}(\bm{z})))] \right\}
\end{equation} 
where $p_{data}$ is the data generating distribution. Both the  $\bm{G}$ and $\bm{D}$ are parametrized networks. Notice that at parameter of $\bm{G}$ fixed, the optimal \textit{discriminator} $\bm{D}^\ast$ has the following expression
\begin{equation}
\bm{D}^\ast(\bm{x}) = \frac{p_{data}(\bm{x})}{p_{data}(\bm{x}) + p_{\bm{G}}(\bm{x})}
\end{equation}
with $p_{\bm{G}}(\bm{x}) = \pi(\bm{z})|\det J_{\bm{G}} |^{-1}$ using the Jacobian of the \textit{generator}, leading to a reformulation of the \texttt{min-max} problem as followed
\begin{equation}
\min_G \left\{ \DKL{\left( p_{data} \left\| \frac{p_{data} + p_{\bm{G}}}{2}\right. \right) }  + \DKL{\left(p_{\bm{G}} \left\| \frac{p_{data} + p_{\bm{G}}}{2} \right. \right)} \right\}
\end{equation}
leading to the optimal solution $p_{\bm{G}} = p_{data}$ yielding a perfect matching of the data distribution.  But, training GAN model via the above objective function is in general unstable with a pathology called \textit{mode-collapse} in the literature, dealing with the fact that the \textit{discriminator} may overfit the dataset too quickly leading to vanishing gradients \citep{Gulrajani2017}. This training instability has been studied by many authors \citep[see e.g.][for some reviews]{Saxena2021,Jozdani2022}. Architectures like \texttt{BigGAN} \citep{Brock2019} and \texttt{StyleGAN2} \cite{Karras2018ASG,Karras2020} overcome the gradient flow issue by different techniques beeing "state-of-the-art" improvements but at the expense of computational resources.

\cite{liu2021towards} manage to address the training stability and low  computational resources, in particular low sized datasets, developing a \texttt{light-weight-GAN} structure. A first ingredient is to use the hinge loss introduced by \cite{Lim2017} which aims to minimize the following losses to alternatively train the \textit{discriminator} and the \textit{generator}: 
\begin{align}
\min_{\bm{D}}& \left\{\Esp_{\bm{x}\sim p_{data}}[\max{(0,1-\bm{D}(\bm{x}))}] + \Esp_{\bm{z}\sim \pi(\bm{z})}[\max{(0,1+\bm{D}(\bm{G}(\bm{z}))}]
\right\}\\
\min_{\bm{G}} &\left\{-\Esp_{\bm{z}\sim \pi(\bm{z})}[\bm{D}(\bm{G}(\bm{z}))]
\right\}
\end{align}
Here $\bm{D}$ is a linear discriminator as in the Support Vector Machine context \citep{Vapnik1997}. The second ingredient concerns a strong regularisation to the \textit{discriminator} loss. The \textit{discriminator}  $\bm{D}$  is treated as an \textit{encoder} (e.g. as in a VAE) and is trained together with \textit{decoders} in such a way that $\bm{D}$ is forced to extract image features at different scales ($8\times 8$, $16\times 16$) that the \textit{decoders} can give good reconstruction. Finally, for the \textit{generator} a new skip-layer excitation module (SLE) is used allowing a more robust gradient flow between feature maps extracted at different scales.

%Coccomini2021 utilise Lightweight GAN
% Schawinski2017,Hemmati_2022 GAN vanilla
GAN architectures have been used for galaxy generation in conjunction of deblending and deconvolving tasks: for instance \cite{Schawinski2017,Hemmati_2022} use vanilla-GANs while  \cite{Coccomini2021} the \texttt{light-weight-GAN}.
%
\subsection{Normalizing Flows}
%
Looking at Equation \ref{eq-VAE-ELBO}, one notices that the optimal solution is reached in case of $\DKL(p_{\bm{\phi}}(\bm{z}|\bm{x})\| p(\bm{z}|\bm{x}))=0$ which is true iif $p_{\bm{\phi}}(\bm{z}|\bm{x}) = p(\bm{z}|\bm{x})$. But this situation cannot be true in general as typical $p_{\bm{\phi}}$ parametrizations involve for instance independent Gaussian distributions. The limited choices of the parametrization is one of the limitation of VAE. So, one motivates the use of normalizing flows (NF) to allow for more flexibility in the distribution family design to hope that the true posterior is member of such family \citep{Tabak2010, Tabak2013a, Rezende2015}.

@@

An answer to the previous question has been given in \cite{kadkhodaie2024generalization} and point out the size of the dataset in the passage from \textit{memorization} to \textit{generation} regimes. We propose to use the same methodology by comparing typically two generative models of same kind trained with two different non-overlapping sub-datasets extracted from the same dataset of original images.

%
\subsection{Datasets}
%
Concerning the training dataset, \cite{kadkhodaie2024generalization}  have used different sources commonly-used in ML as \texttt{CelebA} \citep{Liu2015}, \texttt{LSUN} bedroom \citep{Yu2015} and  \texttt{CelebA HQ} \citep{KarrasALL18} datasets. All dataset images use  gray level single channel ranging in $[0,1]$.  The \texttt{LSUN}  images have been down sampled to $80\times 80$ as well as the \texttt{CelebA} images which also have been further down sampled to $32\times 32$ resolution, while \texttt{CelebA HQ}  have been down sampled to $40\times 40$ resolution.

We are more concerned with galaxy image generation, but as we need to train two models with up to at least $100,000$ images each, we have chosen to use the SDSS (Sloan Digital Sky Survey) dataset described in \citep{smith2021}\footnote{See \url{https://github.com/Smith42/astroddpm} to obtain the raw data.} with some modification on the preprocessing. First, all images are cropped about the target coordinate galaxy to a shape of $64\times 64$ pixels, and secondly from the $(z,g,r)$ channels we produce a single  channel in the range $[0,1]$ using \texttt{scipy} \citep{2020SciPy-NMeth} library\footnote{In particular the \texttt{make\_lupton\_rgb} function has been applied with $Q=8$ and $stretch=0.2$ setting}. Moreover, the gray levels are coded using 5-bit precision in the line of experiment described in \cite{Kingma2018}. With such dataset according to \cite{kadkhodaie2024generalization} results, we expect the transition between memorization to generalisation regimes to occur with a number of training images in the range $10^4-10^5$.

\subsection{Models}
%
{\color{red} Concerning the models, for the GAN model type ....  
For the diffusion-based model, we have used the same \texttt{U-Net} architecture \citep{2015arXiv150504597R}  as \textit{denoiser} consisting of a contracting part and an expending part that allows for multi-scale architecture with scale correlations. Details are found in \cite{kadkhodaie2024generalization}\footnote{Source code:  \url{https://github.com/LabForComputationalVision/memorization_generalization_in_diffusion_models}}. Second, for comparison purposes, we use the \texttt{Glow} flow-based model implemented\footnote{Source code: \url{https://github.com/rosinality/glow-pytorch}} in \texttt{PyTorch}  \citep{PyTorch2019}. The architecture has been presented on Figure \ref{fig-Glow-archi} and we use $K=32$ and $L=4$.}

@@

Figure \ref{fig-Original-Glow-UNet-Gan-samples} presents some galaxy images extracted from the original dataset and generated by  diffusion-based and flow-based models trained with $N=10^5$ original images. 

@@

        \includegraphics[width=0.8\textwidth]{fig-model_sample_images.pdf}
    \caption{Some \textit{validation} images from the original dataset (first row from top) and  generates samples from different models trained with $N=10^5$ images:  the \texttt{light-weight} GAN-based model (second row), the \texttt{Glow} flow-based model (third row) and the diffusion-based with \texttt{U-Net} denoiser (bottow row).}
    \label{fig-Original-Glow-UNet-Gan-samples}

@@

	\caption{Corner plots\protect\footnotemark of morphological coefficients computed with $10^4$ images, first from the original dataset but \textit{never} used by any models (red, "test"), second from a diffusion-based model generation (blue, "diffusion") and from a flow-based model generation (green, "Glow") (both models used for Figure \ref{fig-Original-Glow-UNet-samples}). The figure has been inspired by \cite{HACKSTEIN2023100685}.}

@@

\footnotetext{Plots from the library \url{https://corner.readthedocs.io} \citep{corner2016}.}
